---
title: "05. Markov chain Monte Carlo"
date: "2021-09-26"
output: distill::distill_article
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, dpi = 300, comment = "#>")

library(patchwork)
suppressPackageStartupMessages(library(tidyverse))

theme_set(theme_bw())
```

## Resources

- BDA3 chapter 11 and [reading instructions](../reading-instructions/BDA3_ch11_reading-instructions.pdf)
- lectures:
  - ['5.1. Markov chain Monte Carlo, Gibbs sampling, Metropolis algorithm'](https://aalto.cloud.panopto.eu/Panopto/Pages/Viewer.aspx?id=098dfdb4-f3b8-46aa-b988-aadf00bd3177)
  - ['5.2. Warm-up, convergence diagnostics, R-hat, and effective sample size'](https://aalto.cloud.panopto.eu/Panopto/Pages/Viewer.aspx?id=9f657178-d8cf-4cb8-af62-aadf00cd9423)
- [slides](../slides/slides_ch11.pdf)
- [Assignment 5](assignments/assignment-05.pdf)

## Notes

### Reading instructions

- Outline of the chapter 11
  - Markov chain simulation: before section 11.1, pages 275-276
  - 11.1 Gibbs sampler (an example of simple MCMC method)
  - 11.2 Metropolis and Metropolis-Hastings (an example of simple MCMC method) • 11.3 Using Gibbs and Metropolis as building blocks (can be skipped)
  - 11.4 Inference and assessing convergence (important)
  - 11.5 Effective number of simulation draws (important)
  - 11.6 Example: hierarchical normal model (skip this)
- Animations
  - Nice animations with discussion: http://elevanth.org/blog/2017/11/28/build-a-better-markov-chain/
  - And just the animations with more options to experiment: https://chi-feng.github.io/mcmc-demo/
- Convergence
  - theoretical convergence in an infinite time is different than practical convergence in a finite time
  - no exact moment when chain has converged
    - convergence diagnostics can help to find out if the chain is unlikely to be representative of the target distribution
- $\hat{R}$ effective sample size (ESS, previously $n_\text{eff}$)
  - there are many versions of $\hat{R}$ and effective sample size
    - some software packages compute these using old inferior approaches
    - updated version in [*Rank-normalization, folding, and localization: An improved $\hat{R} for assessing convergence of MCMC*](https://projecteuclid.org/euclid.ba/1593828229)

### Chapter 11. Basics of Markov chain simulation

#### Introduction

- **MCMC**: general method based on drawing values of $\theta$ from approximate distribtuions and then correcting those draws to better approximate the target posterior distribution $p(\theta, y)$
  - **Markov chain**: a sequence of random variables $\theta^1, \theta^2, \dots$ for wihch, for any $t$, the distribtuion of $\theta^t$ given all previous $\theta$'s depends only on the previous value $\theta^{t-1}$
- general process:
  1. create several independent sequences
  2. each sequence $\theta^1, \theta^2, \dots$ starts from some point $\theta^0$
  3. for each $t$, draws $\theta^t$ from the *transition distribution $T_t (\theta^t | \theta^{t-1})$
- essential to check convergence of chains
- this chapter introduces the *Gibbs sampler* and *Metropolic-Hastings algorithm*

#### 11.1 Gibbs sampler

- algorithm:
  1. separate the parameter vector $\theta$ into $d$ components (also called subvectors) $\theta = (\theta_1, \dots, \theta_d)$
  2. for each iteration $t$, each compoent is cycled through (thus, there are $d$ steps for each iteration)
  3. for each iteration $t$, for each $j$ component of $\theta$, each $\theta_j^t$ is sampled from the conditional distribution given all the other current values of $\theta$: $p(\theta_j | \theta_{-j}^{t-1})$
    - where $\theta_{-j}^{t-1} = (\theta_1^t, \dots, \theta_{j-1}^t, \theta_{j+1}^{t-1}, \dots, \theta_d^{t-1})$
    - is just all of the current values of $\theta$ where some have yet to be update in iteration $t$
- ex: bivariate normal distribution
  - a bivariate normally distribution population with mean $\theta = (\theta_1, \theta_2)$ (so $d = 2$ for this example) and covariance matrix $\begin{pmatrix} 1 & \rho \\ \rho  & 1 \\ \end{pmatrix}$
  - given single observation $(y_1, y_2)$
  - uniform prior on $\theta$
  - posterior distribution defined in \@ref(eq:gibbs-posterior)
  - need conditional posterior distribution for each $\theta_j$ on the other components of $\theta$
    - here, use equations A.1 from appendix A (pg. 582): \@ref(eq:gibbs-conditional-posterior)
  - Gibbs sampler just alternatively samples from these two conditional distributions

\begin{equation}
  \begin{pmatrix}
    \theta_1 \\ \theta_2 \end{pmatrix} | y
    \sim \text{N}
    \begin{pmatrix}
    \begin{pmatrix} y_1 \\ y_2 \end{pmatrix},
    \begin{pmatrix} 1 & \rho \\ \rho  & 1 \\ \end{pmatrix}
  \end{pmatrix}
  (\#eq:gibbs-posterior)
\end{equation}

\begin{align}
  \begin{split}
  \theta_1 | \theta_2, y &\sim \text{N}(y_1 + \rho (\theta_2 - y_2), 1 - \rho^2) \\
  \theta_2 | \theta_1, y &\sim \text{N}(y_2 + \rho (\theta_1 - y_1), 1 - \rho^2)
  \end{split}
  (\#eq:gibbs-conditional-posterior)
\end{align}

- below is the code for the example described above

```{r}
chain_to_df <- function(chain, names) {
  purrr::map_dfr(chain, ~ as.data.frame(t(.x))) %>%
    tibble::as_tibble() %>%
    purrr::set_names(names)
}

# Run a single chain of a Gibbs sampler for a bivariate normal distribution.
gibbs_sample_demo <- function(data, rho, theta_t0, N = 100) {
  theta_1 <- theta_t0[[1]]
  theta_2 <- theta_t0[[2]]
  y1 <- data[[1]]
  y2 <- data[[2]]

  chain <- as.list(rep(theta_t0, n = (2 * N) + 1))
  chain[[1]] <- c(theta_1, theta_2, 1)

  for (t in seq(2, N)) {
    theta_1 <- rnorm(1, y1 + rho * (theta_2 - y2), 1 - rho^2)
    chain[[2 * (t - 1)]] <- c(theta_1, theta_2, t)
    theta_2 <- rnorm(1, y2 + rho * (theta_1 - y1), 1 - rho^2)
    chain[[2 * (t - 1) + 1]] <- c(theta_1, theta_2, t)
  }

  chain_df <- chain_to_df(chain, names = c("theta_1", "theta_2", "t"))
  return(chain_df)
}


rho <- 0.8
y <- c(0, 0)
starting_points <- list(
  c(-2.5, -2.5), c(2.5, -2.5), c(-2.5, 2.5), c(2.5, 2.5)
)

set.seed(0)
gibbs_demo_chains <- purrr::map_dfr(
  seq(1, 4),
  ~ gibbs_sample_demo(y, rho, starting_points[[.x]]) %>%
    add_column(chain = as.character(.x))
)

plot_chains <- function(chain_df, x = theta_1, y = theta_2, color = chain) {
  chain_df %>%
    ggplot(aes(x = {{ x }}, y = {{ y }}, color = {{ color }})) +
    geom_path(alpha = 0.6, show.legend = FALSE) +
    scale_color_brewer(type = "qual", palette = "Set1")
}

plot_points <- function(chain_df, x = theta_1, y = theta_2, color = chain) {
  chain_df %>%
    ggplot(aes(x = {{ x }}, y = {{ y }}, color = {{ color }})) +
    geom_point(size = 0.75, alpha = 0.75) +
    scale_color_brewer(type = "qual", palette = "Set1")
}

gibbs_plot_chains <- plot_chains(gibbs_demo_chains)

gibbs_plot_points <- gibbs_demo_chains %>%
  group_by(chain, t) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  plot_points()

gibbs_plot_chains | gibbs_plot_points
```

#### 11.2 Metropolis and Metropolis-Hastings algorithms

- the Metropolis-Hastings algorithm is a generalized version of the Metropolis algorithm

##### The Metropolis algorithm

- is a random walk with an acceptance and rejection rule to converge to the target distribution
- steps:
  1. draws a starting point $\theta^0$ from a *starting distribution* $p_0(\theta)$ such that $p(\theta^0|y) > 0$
  2. for time $t = 1, 2, \dots$:
    a. sample a *proposal* $\theta^*$ from a *jumping/proposal distribution* $J_t(\theta^*|\theta^{t-1})$
    b. calculate the ratio of the densities: $r = \frac{p(\theta^*|y)}{p(\theta^{t-1}|y)}$
    c. set $\theta^t = \theta^*$ with probability $\min(r, 1)$, else $\theta^t = \theta^{t-1}$
- the jumping distribution $J_t$ must be symmetric such that $J_t(\theta_a|\theta_b) = J_t(\theta_b|\theta_a)$
- the iteration still counts even if the proposal $\theta^*$ is rejected
- ex: bivariate normal distribution (same as before):
  - target density as bivariate normal: $p(\theta|y) = \text{N}(\theta | 0, I)$
  - jumping distribution as a bivariate normal with smaller deviations and centered around the previous iteration's $\theta^{t-1}$: $J_t(\theta^*|\theta^{t-1}) = \text{N}(\theta^* | \theta^{t-1}, 0.2^2I)$
  - thus, the density ratio: $r = \text{N}(\theta^*|0, I) / \text{N}(\theta^{t-1}|0, I)$

**TODO: fix the issue below.**

```{r}
calc_metropolis_density_ratio <- function(t_star, t_m1, data, prior_cov_mat, jumping_cov_mat) {
  numerator <- mvtnorm::dmvnorm(t_star, data, prior_cov_mat)
  denominator <- mvtnorm::dmvnorm(t_m1, data, jumping_cov_mat)
  return(numerator / denominator)
}

metropolis_algorithm_demo <- function(data, theta_t0, N = 1000) {
  theta_t <- unlist(theta_t0)
  prior_dist_mu <- data
  prior_dist_cov_mat <- matrix(c(1, 0, 0, 1), nrow = 2)
  jumping_dist_cov_mat <- prior_dist_cov_mat * 0.2^2

  chain <- as.list(rep(NA_real_, n = N + 1))
  chain[[1]] <- theta_t

  for (t in seq(2, N + 1)) {
    theta_star <- mvtnorm::rmvnorm(
      n = 1, mean = theta_t, sigma = jumping_dist_cov_mat
    )[1, ]

    density_ratio <- calc_metropolis_density_ratio(
      t_star = theta_star,
      t_m1 = theta_t,
      data = data,
      prior_cov_mat = prior_dist_cov_mat,
      jumping_cov_mat = jumping_dist_cov_mat
    )

    accept <- runif(1) < min(c(1, density_ratio))
    if (accept) {
      theta_t <- theta_star
    }
    chain[[t]] <- theta_t
  }
  return(chain_to_df(chain, names = c("theta_1", "theta_2")))
}

set.seed(0)
metropolis_chains <- purrr::map_dfr(
  seq(1, 4),
  ~ metropolis_algorithm_demo(c(0, 0), starting_points[[.x]]) %>%
    add_column(chain = as.character(.x))
)

metropolis_plot_chains <- plot_chains(metropolis_chains)
metropolis_plot_points <- metropolis_chains %>%
  group_by(chain) %>%
  slice_tail(n = 500) %>%
  plot_points()
metropolis_plot_chains | metropolis_plot_points
```



### Lecture notes
