---
title: "Assignment 2"
date: "2021-08-30"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", dpi = 300)
```

**[Assignment 2](assignment-02.pdf)**

## Setup

```{r, warning=FALSE, message=FALSE}
# Libraries
library(aaltobda)
library(glue)
library(tidyverse)

# Data
algae <- readLines(here::here("data", "algae.txt"))
algae <- as.integer(algae)
```


## Exercise 1. Inference for binomial proportion

**Algae status is monitored in 274 sites at Finnish lakes and rivers.**
**The observations for the 2008 algae status at each site are presented in file "algae.txt" (’0’: no algae, ’1’: algae present).**
**Let $\pi$ be the probability of a monitoring site having detectable blue-green algae levels and $y$ the observations in algae.**
**Use a binomial model for the observations $y$ and a $\text{Beta}(2, 10)$ prior for binomial model parameter $\pi$ to formulate a Bayesian model.**

**a) Formulate (1) the likelihood $p(y|\pi)$ as a function of $\pi$, (2) the prior $p(\pi)$, and (3) the resulting posterior $p(\pi|y)$.**

```{r}
print(head(algae))
print(paste("Number of data points:", length(algae)))
print(paste("Number of 1's:", sum(algae)))
print(paste("Prop. of 1's:", round(mean(algae), 3)))
```


1. likelihood: $p(y|\pi) = \text{Beta}(44, 274-44)$
2. prior: $p(\pi) = \text{Beta}(2, 10)$
3. posterior: $p(\pi|y) = \text{Beta}(46, 240)$

**b) What can you say about the value of the unknown \pi according to the observations and your prior knowledge?**
**Summarize your results with a point estimate (i.e. $E(\pi|y)$) and a 90% posterior interval.**

Some test data provided by the question to check if I'm on the right track.

```{r}
algae_test <- c(0, 1, 1, 0, 0, 0)
```

```{r}
beta_point_est <- function(prior_alpha, prior_beta, data) {
  y <- sum(data)
  n <- length(data)
  posterior <- (prior_alpha + y) / (prior_alpha + prior_beta + n)
  return(posterior)
}

posterior_test <- beta_point_est(
  prior_alpha = 2, prior_beta = 10, data = algae_test
)
stopifnot(round(posterior_test, 7) == 0.2222222)

beta_point_est(prior_alpha = 2, prior_beta = 10, data = algae)
```

```{r}
beta_interval <- function(prior_alpha, prior_beta, data, prob = 0.9) {
  y <- sum(data)
  n <- length(data)
  p_low <- (1 - prob) / 2
  q_low <- qbeta(p_low, prior_alpha + y, prior_beta + n - y)
  q_high <- qbeta(1 - p_low, prior_alpha + y, prior_beta + n - y)
  return(c(q_low, q_high))
}

posterior_interval_test <- beta_interval(
  prior_alpha = 2, prior_beta = 10, data = algae_test, prob = 0.9
)
stopifnot(round(posterior_interval_test, 7) == c(0.0846451, 0.3956414))

beta_interval(prior_alpha = 2, prior_beta = 10, data = algae, prob = 0.9)
```

**c) What is the probability that the proportion of monitoring sites with detectable algae levels $\pi$ is smaller than $\pi_\theta = 0.2$ that is known from historical records?**

```{r}
beta_low <- function(prior_alpha, prior_beta, data, pi_0 = 0.2) {
  y <- sum(data)
  n <- length(data)
  prob <- pbeta(pi_0, prior_alpha + y, prior_beta + n - y, lower.tail = TRUE)
  return(prob)
}

test_prob <- beta_low(
  prior_alpha = 2, prior_beta = 10, data = algae_test, pi_0 = 0.2
)
stopifnot(round(test_prob, 7) == 0.4511238)

beta_low(prior_alpha = 2, prior_beta = 10, data = algae, pi_0 = 0.2)
```

**d) What assumptions are required in order to use this kind of a model with this type of data?**
**(No need to discuss exchangeability yet, as it is discussed in more detail in BDA Chapter 5 and Lecture 7.)**

We need to assume that the data is independently and identically distributed, which included the assumption that the data is exchangable.
We are also assuming the data is accuractely collected in a consistent manner.
We are assuming there are no subgroups within the data that would necessitate a hierarchical model to account for the random effects variation.

**e) Make prior sensitivity analysis by testing a couple of different reasonable priors and plot the different posteriors.**
**Summarize the results by one or two sentences.**

```{r}
# Some interesting priors.
priors <- tibble::tribble(
  ~prior_alpha, ~prior_beta,
  1, 1,
  2, 2,
  2, 10,
  4, 20,
  20, 100,
) %>%
  mutate(
    lbl = glue("Beta({prior_alpha},{prior_beta})"),
    lbl = fct_inorder(lbl)
  )
priors
```

```{r}
posterior_distribution <- function(prior_alpha,
                                   prior_beta,
                                   data,
                                   step = 0.001,
                                   ...) {
  y <- sum(data)
  n <- length(data)
  pi <- seq(0, 1, step)
  posterior <- dbeta(pi, prior_alpha + y, prior_beta + n - y)
  return(tibble(pi = pi, posterior = posterior))
}

# Get the posterior distribution for each prior.
posteriors <- priors %>%
  mutate(posterior = purrr::map2(
    prior_alpha, prior_beta, posterior_distribution,
    data = algae
  )) %>%
  unnest(posterior)

# Plot the most interesting region of the posteriors.
posteriors %>%
  filter(0.05 < pi & pi < 0.3) %>%
  ggplot(aes(x = pi, y = posterior)) +
  geom_line(aes(group = lbl, color = lbl), size = 0.9) +
  scale_x_continuous(expand = expansion()) +
  scale_y_continuous(expand = expansion(mult = c(0.01, 0.02))) +
  scale_color_brewer(type = "qual", palette = "Set1") +
  theme_bw() +
  theme(legend.position = c(0.85, 0.7)) +
  labs(x = "pi", y = "posterior probability", color = "prior")
```

The posterior is not very sensitive to the prior save for the exception of an overly-confident prior of $\text{Beta}(20, 100)$.
This is likely due to the large amount of data, meaning that the likelihood dominated the posterior.
